Portal Forecast

### 2018-12-03 evaluation.Rmd
Reorganize, cleanup function writing

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE
)
library(dplyr)
library(portalr)
library(yaml)
source("tools/forecast_tools.R")

model_metadata = yaml.load_file("data/model_metadata.yaml")
pred_interval = model_metadata$pred_interval
models_of_interest = c('AutoArima','Ensemble','ESSS','nbGARCH','pevGARCH')

species_of_interest = c('BA','DM','DO','PP','OT','NA','total')
species_names = portalr:::full_path('PortalData/Rodents/Portal_rodent_species.csv', '~') %>%
  read.csv(stringsAsFactors=FALSE, na.strings = '') %>%
  select(species = speciescode, full_species_name = scientificname)

#add a total entry
species_names = species_names %>%
  add_row(species='total', full_species_name='Total Rodents')

new_moon_file = portalr:::full_path('PortalData/Rodents/moon_dates.csv', '~')
new_moons = read.csv(new_moon_file)
new_moons$censusdate = as.Date(new_moons$censusdate)

# Setup rodent observations
species_abundance = portalr::abundance("repo", clean=FALSE, shape='flat', level='treatment', 
                                       length="longterm", time='newmoon') %>%
  filter(treatment=="control") %>%
  rename(actual = abundance, level = treatment) %>%
  mutate(level='Controls',currency='abundance') %>%
  filter(species %in% species_of_interest)

total_abundance = portalr::abundance("repo", clean=FALSE, shape='flat', level='treatment', 
                                     length="longterm", time='newmoon') %>%
  filter(treatment=="control") %>%
  group_by(newmoonnumber) %>%
  summarise(actual=sum(abundance)) %>%
  ungroup() %>%
  mutate(level='Controls',currency='abundance',species='total')

observation_data = species_abundance %>%
  bind_rows(total_abundance) %>% 
  left_join(new_moons, by='newmoonnumber')

@@

This graph shows the forecast error of some recent forecasts for the most recent sampling dates. This is shown for all model types in the columns and several common rodent species (and total rodents) in the rows. These evaluations are for the Control plots where no manipulation is done. 
```{r recent_eval, echo=FALSE, message=FALSE, warning=FALSE,, fig.width=15, fig.height=15}
###############################################
#Get all the recent forecasts
forecast_data = compile_forecasts(use_hindcasts = FALSE) %>%
  filter(model %in% models_of_interest)

@@

# Keep 1 forecast per unique initial_newmoon. When there are multiple forecasts
# using the same initial_newmoon, use the first one.
forecast_dates_to_keep = forecast_data %>%
  select(date, initial_newmoon) %>%
  distinct() %>%
  group_by(initial_newmoon) %>%
  filter(date == min(date)) %>%
  ungroup() %>%
  mutate(keep='yes')

forecast_data = forecast_data %>%
  left_join(forecast_dates_to_keep, by=c('date','initial_newmoon')) %>%
  filter(keep=='yes') %>%
  select(-keep)

##################################
#Combine forecasts and oservations
forecast_errors = forecast_data %>%
  left_join(observation_data, by=c('newmoonnumber','level','species','currency')) %>% 
  mutate(rmse = sqrt((estimate - actual)^2))

# Sometimes there are model runs on the same day and with all the same info,
# this gets rid of those
forecast_errors = forecast_errors %>%
  distinct()

# Drop any entries that don't have an observation
# (ie. a rainy sample period)
forecast_errors = forecast_errors %>%
  filter(!is.na(actual))

# Filter to common species and apply the full name
forecast_errors = forecast_errors %>%
  filter(species %in% species_of_interest) %>%
  left_join(species_names, by='species') %>%
  select(-species) %>%
  rename(species=full_species_name)

@@

#############################################################
# Displaying the data from many forecasts in the past can create a mess of a plot. see #218.
# So here three forecasts are chosen from 1, 6, and 12 newmoons into the past. 
# Sometimes forecasts data is missing for a particular initial_newmoon, the recursive function here
# attempts to find another one from the same time pluts/minus 2 month period. This ensures 3 forecasts
# are always shown.

# If there is no data for initial_newmoon 500, this will check nearby dates in this order: 499,501,498,402
update_initial_newmoon = function(potential_initial_newmoon, attempt=1){
  if(attempt==6){
    return(NA)
  }
  lead_time_data = forecast_errors %>%
    filter(initial_newmoon == potential_initial_newmoon)

  # Alternate steps of -1, +2, -3, +4
  next_try_step = ifelse(attempt%%2 == 0, attempt, attempt * -1)

  # no data to show? then try one month back
  if(nrow(lead_time_data)==0){
    #print(paste('Try ',attempt, ', Nothing with ',potential_initial_newmoon,' trying ',potential_initial_newmoon+next_try_step))
    return(update_initial_newmoon(potential_initial_newmoon+next_try_step, attempt+1))
  } else {
    return(potential_initial_newmoon)
  }

}
# Adjust this to add more than 3 forecast dates, or change the lead times
lead_times_to_show = c(1,6,12)
most_recent_observed_newmoon = max(new_moons$newmoonnumber)
initial_newmoons_to_keep = most_recent_observed_newmoon - lead_times_to_show

@@

forecast_errors = forecast_errors %>%
  filter(initial_newmoon %in% initial_newmoons_to_keep)

#####################################
ggplot(forecast_errors, aes(x=censusdate, y=rmse, group=as.character(date), color=as.character(date))) +
  geom_point(size=4)+
  geom_line(size=1.5) +
  scale_color_brewer(palette='Dark2') +
  scale_x_date(breaks = unique(forecast_errors$censusdate)) +
  facet_grid(species~model, scales = 'free_y') + 
  theme_bw() +
  theme(axis.text.x = element_text(angle=45, hjust=0.99, debug = FALSE),
        axis.text = element_text(size=12),
        axis.title = element_text(size=20),
        strip.text = element_text(size=12),
        legend.text = element_text(size=15),
        legend.title = element_text(size=18)) +
  labs(x='Sample Date',y='RMSE', color='Forecast Issue\nDate')

@@

#Get the all the forecasts made during observation period
forecast_data = compile_forecasts(use_hindcasts = TRUE)
forecast_errors = calculate_forecast_error(observation_data, forecast_data, error_metric = 'RMSE') %>%
  filter(error_value < 200) %>% #Drop RMSE greater than this because it throws off all the graphs
  bind_rows(calculate_forecast_error(observation_data, forecast_data, error_metric = 'coverage'))

forecast_errors = forecast_errors %>%
  left_join(species_names, by='species')

ggplot(forecast_errors, aes(x=lead_time, y=error_value, group=model, color=model)) +
  geom_point()+
  geom_line() +
  geom_hline(yintercept = 0.9) +
  labs(x='Lead Time (New Moons)') +
  facet_wrap(full_species_name~error_metric, scales = 'free_y', ncol=2) + 
  theme_bw() +
  labs(y = "Error Value", colour = "Model")

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE
)
'%>%' <- magrittr::'%>%'
source("tools/forecast_tools.R")

model_metadata = yaml::yaml.load_file("data/metadata.yaml")
pred_interval = model_metadata$pred_interval
models_of_interest = c('AutoArima','Ensemble','ESSS','nbGARCH','pevGARCH')

species_of_interest = c('BA','DM','DO','PP','OT','NA','total')
species_names = read.csv('PortalData/Rodents/Portal_rodent_species.csv',
                         stringsAsFactors=FALSE, na.strings = '') %>%
  dplyr::select(species = speciescode, full_species_name = scientificname) %>%
  dplyr::add_row(species='total', full_species_name='Total Rodents')

new_moons = read.csv('PortalData/Rodents/moon_dates.csv') %>%
            dplyr::mutate(censusdate = as.Date(censusdate))

# Setup rodent observations
species_abundance = portalr::abundance(clean=FALSE, shape='flat', level='treatment', 
                                       plots="longterm", time='newmoon') %>%
  dplyr::filter(treatment=="control") %>%
  dplyr::rename(actual = abundance, level = treatment) %>%
  dplyr::mutate(level='Controls',currency='abundance') %>%
  dplyr::filter(species %in% species_of_interest)

total_abundance = portalr::abundance(clean=FALSE, shape='flat', level='treatment', 
                                     plots="longterm", time='newmoon') %>%
  dplyr::filter(treatment=="control") %>%
  dplyr::group_by(newmoonnumber) %>%
  dplyr::summarise(actual=sum(abundance)) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(level='Controls',currency='abundance',species='total')

observation_data = species_abundance %>%
  dplyr::bind_rows(total_abundance) %>% 
  dplyr::left_join(new_moons, by='newmoonnumber')

@@

This graph shows the forecast error of some recent forecasts for the most recent sampling dates. This is shown for all model types in the columns and several common rodent species (and total rodents) in the rows. These evaluations are for the Control plots where no manipulation is done. 
```{r recent_eval, echo=FALSE, message=FALSE, warning=FALSE,, fig.width=15, fig.height=15}
###############################################
#Get all the recent forecasts
forecast_data = compile_forecasts(use_hindcasts = FALSE) %>%
  dplyr::filter(model %in% models_of_interest)

@@

# Keep 1 forecast per unique initial_newmoon. When there are multiple forecasts
# using the same initial_newmoon, use the first one.
forecast_dates_to_keep = forecast_data %>%
  dplyr::select(date, initial_newmoon) %>%
  dplyr::distinct() %>%
  dplyr::group_by(initial_newmoon) %>%
  dplyr::filter(date == min(date)) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(keep='yes')

forecast_data = forecast_data %>%
  dplyr::left_join(forecast_dates_to_keep, by=c('date','initial_newmoon')) %>%
  dplyr::filter(keep=='yes') %>%
  dplyr::select(-keep)

##################################
#Combine forecasts and oservations
forecast_errors = forecast_data %>%
  dplyr::left_join(observation_data, by=c('newmoonnumber','level','species','currency')) %>% 
  dplyr::mutate(rmse = sqrt((estimate - actual)^2))

# Sometimes there are model runs on the same day and with all the same info,
# this gets rid of those
forecast_errors = forecast_errors %>%
  dplyr::distinct()

# Drop any entries that don't have an observation
# (ie. a rainy sample period)
forecast_errors = forecast_errors %>%
  dplyr::filter(!is.na(actual))

# Filter to common species and apply the full name
forecast_errors = forecast_errors %>%
  dplyr::filter(species %in% species_of_interest) %>%
  dplyr::left_join(species_names, by='species') %>%
  dplyr::select(-species) %>%
  dplyr::rename(species=full_species_name)

@@

#############################################################
# Displaying the data from many forecasts in the past can create a mess of a plot. see #218.
# So here three forecasts are chosen from 1, 6, and 12 newmoons into the past. 
# Sometimes forecasts data is missing for a particular initial_newmoon, the recursive function 
# update_initial_newmoon attempts to find another one from the same time pluts/minus 2 month period.
#  This ensures 3 forecasts are always shown.
lead_times_to_show = c(1,6,12)
most_recent_observed_newmoon = max(new_moons$newmoonnumber)
initial_newmoons_to_keep = most_recent_observed_newmoon - lead_times_to_show

@@

forecast_errors = forecast_errors %>%
  dplyr::filter(initial_newmoon %in% initial_newmoons_to_keep)

#####################################
ggplot2::ggplot(forecast_errors, ggplot2::aes(x=censusdate, y=rmse, group=as.character(date), color=as.character(date))) +
  ggplot2::geom_point(size=4)+
  ggplot2::geom_line(size=1.5) +
  ggplot2::scale_color_brewer(palette='Dark2') +
  ggplot2::scale_x_date(breaks = unique(forecast_errors$censusdate)) +
  ggplot2::facet_grid(species~model, scales = 'free_y') + 
  ggplot2::theme_bw() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle=45, hjust=0.99, debug = FALSE),
        axis.text = ggplot2::element_text(size=12),
        axis.title = ggplot2::element_text(size=20),
        strip.text = ggplot2::element_text(size=12),
        legend.text = ggplot2::element_text(size=15),
        legend.title = ggplot2::element_text(size=18)) +
  ggplot2::labs(x='Sample Date',y='RMSE', color='Forecast Issue\nDate')

@@

#Get the all the forecasts made during observation period
forecast_data = compile_forecasts(use_hindcasts = TRUE)
forecast_errors = calculate_forecast_error(observation_data, forecast_data, error_metric = 'RMSE') %>%
  dplyr::filter(error_value < 200) %>% #Drop RMSE greater than this because it throws off all the graphs
  dplyr::bind_rows(calculate_forecast_error(observation_data, forecast_data, error_metric = 'coverage'))

forecast_errors = forecast_errors %>%
  dplyr::left_join(species_names, by='species')

ggplot2::ggplot(forecast_errors, ggplot2::aes(x=lead_time, y=error_value, group=model, color=model)) +
  ggplot2::geom_point()+
  ggplot2::geom_line() +
  ggplot2::geom_hline(yintercept = 0.9) +
  ggplot2::labs(x='Lead Time (New Moons)') +
  ggplot2::facet_wrap(full_species_name~error_metric, scales = 'free_y', ncol=2) + 
  ggplot2::theme_bw() +
  ggplot2::labs(y = "Error Value", colour = "Model")

### 2018-12-03 index.Rmd
Reorganize, cleanup function writing

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(portalr)
source("tools/forecast_tools.R")
source("tools/data_tools.R")
```

@@

```{r, echo=FALSE, message=FALSE, warning=FALSE}
obs_data = portalr::abundance("repo", clean=FALSE, level='treatment', length="longterm") %>%
  filter(treatment=="control")
obs_data$total = rowSums(select(obs_data, -c(period, treatment)))
new_moons = read.csv(text = RCurl::getURL("https://raw.githubusercontent.com/weecology/PortalData/master/Rodents/moon_dates.csv"))

obs_data_newmoon = inner_join(obs_data, new_moons, by = c("period" = "period"))
obs_data_newmoon$censusdate = as.Date(obs_data_newmoon$censusdate)

@@

most_recent_forecast = max(for_data$date)
for_data = filter(for_data, date == most_recent_forecast)

for_data$model = "ensemble"
# get dates for future new moons for plotting
future_moons = get_moon_data() %>% get_future_moons() %>% select(c("newmoonnumber","newmoondate"))
for_data = inner_join(for_data, future_moons, by = c("newmoonnumber" = "newmoonnumber"))

forecast_viz(obs_data = obs_data_newmoon,
             obs_date_col_name = "censusdate",
             obs_val_col_name = "total",
             for_data = for_data,
             for_date_col_name = "newmoondate",
             for_val_col_name = "estimate",
             for_model_name = "ensemble",
             for_lowerpi_col_name = "LowerPI",
             for_upperpi_col_name = "UpperPI",
             start_newmoon = 300,
             ylabel = 'Total Abundance')
```

@@

```{r speciesforecasts, echo=FALSE, message=FALSE, warning=FALSE}
source("tools/forecast_tools.R")

data = compile_forecasts()
ensemble = dplyr::filter(data, level == 'Controls', model == 'Ensemble', date == max(date))
sp_predictions = get_sp_predicts(ensemble, 'Controls', 
            lead_time = length(unique(ensemble$newmoonnumber)) - 11)
sp_predict = plot_species_forecast(sp_predictions,
            title=paste0(sp_predictions$forecast_date[2], ": Control plots"))

@@

This figures are similar to the primary Total Abundance forecast above, but for the abundance of three common rodent species. 
```{r highabund_ts_forecasts, echo=FALSE, message=FALSE, warning=FALSE}
most_abund_sp = sp_predictions %>% 
  filter(species != 'total') %>% 
  arrange(desc(estimate)) %>% 
  head(3) %>% 
  select(species)

# load in rodent species table to get scientific names to display on plots
species_table = read.csv(
  text = RCurl::getURL(
      "https://raw.githubusercontent.com/weecology/PortalData/master/Rodents/Portal_rodent_species.csv"),stringsAsFactors = F,na.strings = '')
most_abund_sp_names = filter(species_table,speciescode %in% most_abund_sp$species) %>% select(speciescode,scientificname)

for (n in seq(dim(most_abund_sp_names)[1])) {
  species_forecast = forecast_viz(obs_data = obs_data_newmoon,
               obs_date_col_name = "censusdate",
               obs_val_col_name = most_abund_sp_names$speciescode[n],
               for_data = for_data,
               for_date_col_name = "newmoondate",
               for_val_col_name = "estimate",
               for_model_name = "ensemble",
               for_lowerpi_col_name = "LowerPI",
               for_upperpi_col_name = "UpperPI",
               start_newmoon = 300,
               ylabel = most_abund_sp_names$scientificname[n])
  plot(species_forecast)
}
```

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
'%>%' <- magrittr::'%>%'
source("tools/forecast_tools.R")
```

@@

```{r, echo=FALSE, message=FALSE, warning=FALSE}
obs_data = portalr::abundance(clean=FALSE, level='treatment', plots="longterm") %>%
  dplyr::filter(treatment=="control")
obs_data$total = rowSums(dplyr::select(obs_data, -c(period, treatment)))
new_moons = read.csv('PortalData/Rodents/moon_dates.csv') %>%
                   dplyr::mutate(censusdate = as.Date(censusdate),
                          year = lubridate::year(newmoondate),
                          month = lubridate::month(newmoondate))

obs_data_newmoon = dplyr::inner_join(obs_data, new_moons, by = c("period" = "period"))

@@

# get dates for future new moons for plotting
future_moons = new_moons %>% portalr::get_future_moons() %>%    
                         dplyr::select(c("newmoonnumber","newmoondate"))
for_data = dplyr::inner_join(for_data, future_moons, by = c("newmoonnumber" = "newmoonnumber"))

forecast_viz(obs_data = obs_data_newmoon,
             obs_date_col_name = "censusdate",
             obs_val_col_name = "total",
             for_data = for_data,
             for_date_col_name = "newmoondate",
             for_val_col_name = "estimate",
             for_model_name = "Ensemble",
             for_lowerpi_col_name = "LowerPI",
             for_upperpi_col_name = "UpperPI",
             start_newmoon = 300,
             ylabel = 'Total Abundance')
```

@@

```{r speciesforecasts, echo=FALSE, message=FALSE, warning=FALSE}

sp_predictions = get_sp_predicts(for_data, 'Controls', 
            lead_time = length(unique(for_data$newmoonnumber)) - 11)
sp_predict = plot_species_forecast(sp_predictions,
            title=paste0(sp_predictions$forecast_date[2], ": Control plots"))

@@

This figures are similar to the primary Total Abundance forecast above, but for the abundance of three common rodent species. 
```{r highabund_ts_forecasts, echo=FALSE, message=FALSE, warning=FALSE}
most_abund_sp = sp_predictions %>% 
  dplyr::filter(species != 'total') %>% 
  dplyr::arrange(desc(estimate)) %>% 
  head(3) %>% 
  dplyr::select(species)

# load in rodent species table to get scientific names to display on plots
species_table = read.csv('PortalData/Rodents/Portal_rodent_species.csv',
                          stringsAsFactors = F,na.strings = '')
most_abund_sp_names = dplyr::filter(species_table,speciescode %in% most_abund_sp$species) %>%      
                      dplyr::select(speciescode,scientificname)

for (n in seq(dim(most_abund_sp_names)[1])) {
  species_forecast = forecast_viz(obs_data = obs_data_newmoon,
               obs_date_col_name = "censusdate",
               obs_val_col_name = most_abund_sp_names$speciescode[n],
               for_data = for_data,
               for_date_col_name = "newmoondate",
               for_val_col_name = "estimate",
               for_model_name = "Ensemble",
               for_lowerpi_col_name = "LowerPI",
               for_upperpi_col_name = "UpperPI",
               start_newmoon = 300,
               ylabel = most_abund_sp_names$scientificname[n])
  plot(species_forecast)
}
```
